### 2022-10-04 NSF2022/Description/background.tex

Motivated by its broad applicability to biology, chemistry, and engineering,
PIs RR and YNY developed a mathematical model, called the
hydrophobic attraction potential (HAP) model~\cite{Fu2018_SIAM},
that is based on
the linear response of water to surface perturbations. This model was
used by all the PIs to show how amphiphilic Janus particles (JP)
self-assemble into a vesicle-like structure~\cite{FuQuRyYo20}.
Janus particles are typically spherical particles with a biphasic
material label on either hemisphere \cite{CaFaRaVe89,Gaetal13,Mallory2017,HaBr20,McBr21}.
Based on
preliminary results (\S\ref{sec:preliminary_work}), the PIs propose to
extend this HAP model as a general methodology that leads
to new mathematical ideas and efficient numerical methods.

@@

The HAP is a conceptual advance for modeling bilayer membranes.
Continuum elastic energies of membrane assume an energy density that
depends quadratically on the bilayer strain
\cite{Hamm2000,TerziDeserno17, PhysRevE.102.042406}.  The coefficients
of the continuum ansatz correspond to experimentally measured elastic
moduli \cite{Nagle17, Nagle17-2, NAGLE2000159}.  The HAP approach can
potentially explain membrane properties from first principles because it
derives elastic energies assuming parameters for the water structure and
particle shape \cite{Fu2018_SIAM, FuQuRyYo20, doi:10.1063/5.0009734,
LiAn-Chang16}.  Immersed boundary and boundary integral are highly
effective for modeling intact, giant unilamellar vesicles (GUV) because
they use relatively large time-step sizes and mesh sizes
\cite{
Shravan09,
Rahimian15,
KimLai2010_JCP,
KimLai2012_PRE,
HuLaiSeolEtAl2016_JCP,
Bartels,
Peng13,
RyKlYaCo16,
Sinha15,
Lowengrub13,
Hu,
Hu13}.
Conversely, the number of coarse-grained particles needed to track a GUV
in the HAP formalism is relatively large.  For microscopic process like
fusion, however, membrane deformations occur at a length scale
comparable to the thickness of the membrane
\cite{Kuzmin7235,Aeffner2012,KoKo2002}.  The small deformation
assumption of continuum theory no longer holds
\cite{Hamm2000,TerziDeserno17, PhysRevE.102.042406}.  The HAP theory
does encode membrane thickness through particle diameter, and it is
comparatively straightforward to account for proteins and lipid mixtures
by including several particle species.  Any ruptures or insertions occur
self-consistently in the HAP formulation.  Phase-field and level-set
methods also allow for topological changes but they require multiple
phase fields and level sets and additional (sometimes empirical) laws to
couple them to model membrane-membrane interactions
\cite{
DuLiuWang2004_JCP,
BibenKassnerMisbah2005_PRE,
DoyeuxGuyotChabannesEtAl2013_JCAM,
Du05,
QiangDu08,
doi:10.1098/rspa.2012.0505,
doi:10.1137/130941432,
Feetzl18,
doi:10.1137/16M1108406}.
In addition, these methods require the entire fluid and interfacial
domains to be discretized.  In contrast, the HAP methodology allows
evolution equations to be solved using integral equation methods that
when optimized have linear or near-linear computational complexity in
the number of particles~\cite{fmm1, fmm2, fmm3, fmm4, fmm5, fmm6, fmm7,
fmm8}.

>>>>

Motivated by its broad applicability to biology, chemistry, and engineering,
PIs RR and YNY developed a mathematical model, called the
hydrophobic attraction potential (HAP) model~\cite{Fu2018_SIAM},
that is based on
the linear response of water to surface perturbations. This model was
used by all the PIs to show how amphiphilic Janus particles (JP)
self-assemble into a vesicle-like structure~\cite{FuQuRyYo22}.
Janus particles are typically spherical particles with a biphasic
material label on either hemisphere \cite{CaFaRaVe89,Gaetal13,Mallory2017,HaBr20,McBr21}.
Based on
preliminary results (\S\ref{sec:preliminary_work}), the PIs propose to
extend this HAP model as a general methodology that leads
to new mathematical ideas and efficient numerical methods.

@@

The HAP is a conceptual advance for modeling bilayer membranes.
Continuum elastic energies of membrane assume an energy density that
depends quadratically on the bilayer strain
\cite{Hamm2000,TerziDeserno17, PhysRevE.102.042406}.  The coefficients
of the continuum ansatz correspond to experimentally measured elastic
moduli \cite{Nagle17, Nagle17-2, NAGLE2000159}.  The HAP approach can
potentially explain membrane properties from first principles because it
derives elastic energies assuming parameters for the water structure and
particle shape \cite{Fu2018_SIAM, FuQuRyYo22, doi:10.1063/5.0009734,
LiAn-Chang16}.  Immersed boundary and boundary integral are highly
effective for modeling intact, giant unilamellar vesicles (GUV) because
they use relatively large time-step sizes and mesh sizes
\cite{
Shravan09,
Rahimian15,
KimLai2010_JCP,
KimLai2012_PRE,
HuLaiSeolEtAl2016_JCP,
Bartels,
Peng13,
RyKlYaCo16,
Sinha15,
Lowengrub13,
Hu,
Hu13}.
Conversely, the number of coarse-grained particles needed to track a GUV
in the HAP formalism is relatively large.  For microscopic process like
fusion, however, membrane deformations occur at a length scale
comparable to the thickness of the membrane
\cite{Kuzmin7235,Aeffner2012,KoKo2002}.  The small deformation
assumption of continuum theory no longer holds
\cite{Hamm2000,TerziDeserno17, PhysRevE.102.042406}.  The HAP theory
does encode membrane thickness through particle diameter, and it is
comparatively straightforward to account for proteins and lipid mixtures
by including several particle species.  Any ruptures or insertions occur
self-consistently in the HAP formulation.  Phase-field and level-set
methods also allow for topological changes but they require multiple
phase fields and level sets and additional (sometimes empirical) laws to
couple them to model membrane-membrane interactions
\cite{
DuLiuWang2004_JCP,
BibenKassnerMisbah2005_PRE,
DoyeuxGuyotChabannesEtAl2013_JCAM,
Du05,
QiangDu08,
doi:10.1098/rspa.2012.0505,
doi:10.1137/130941432,
Feetzl18,
doi:10.1137/16M1108406}.
In addition, these methods require the entire fluid and interfacial
domains to be discretized.  In contrast, the HAP methodology allows
evolution equations to be solved using integral equation methods that
when optimized have linear or near-linear computational complexity in
the number of particles~\cite{fmm1, fmm2, fmm3, fmm4, fmm5, fmm6, fmm7,
fmm8}.

### 2022-10-04 NSF2022/Description/specificaim3.tex

\subsection{Specific Aim 3: Efficient, high-order methods for
large-scale simulations}
\label{subsec:specific_aim_3}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 2, we propose novel and efficient high-order numerical methods to
simulate dynamic assembly of particles under the HAP in a viscous
solvent. As described in~\eqref{eq:stokes} and illustrated in
Figure~\ref{fig:flow_map}, the solvent phase is modeled by the Stokes
equations for an incompressible, viscous fluid, and these equations are
coupled to the screened-Laplace equation~\eqref{eq:SL} through viscous
and hydrophobic stress balance.

The hydrophobic interactions come from the tendency of particles to
minimize the free energy of the structure of the surrounding water
molecules. The HAP model uses the free energy
\begin{align}
\label{eq:free_energy}
F[u] = C \int_{\Omega} \rho |\nabla u|^2 + \rho^{-1} f(u) \,dx
\end{align}
where $u$ is an order parameter for the structure of water, $\rho$ is a
decay length, $C$ is a constant, and $f(u)$ is a potential.
Minimizing $F[u]$ requires that $u$ satisfies 
\begin{align}
  \label{eq:SL}
  -\rho^2 \Delta u + f'(u) = 0  \text{ in } \Omega,\quad u = g
  \text{ on } \Omega.
\end{align}
where the boundary condition $g$ is a material label that is transported
with the particle motions.

The HAP model requires solving exterior problems for the order
parameter~\eqref{eq:SL} and the Stokes equations~\eqref{eq:stokes}. We
will develop methods based on combinations of volume and layer
potentials, and inexact Newton's method. Numerical methods to solve
these integral equations (IEs) have several advantages over their
PDE-based counterpart: by construction, they satisfy far-field and mass
conservation conditions; only $\partial\Omega$ needs to be discretized
for layer potentials; carefully chosen IEs result in a well-conditioned
linear system that is solved with a mesh-independent number of GMRES
iterations; and high-order or spectral accuracy is attainable with
appropriate quadrature methods. The PIs have demonstrated that IEs are a
powerful tool to simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo20}. Moreover, elements of these
formulations have been described in three dimensions~\cite{ying_2006,
manasthesis, rac-gre2016}. This proposal will extend the two-dimensional
methods to more general amphiphilic particle models that
satisfy~\eqref{eq:SL}, and all the methods can naturally be extended to
three dimensions. These methods will be motivated by models and analysis
developed in Specific Aims 1 and 3. \S\ref{subsec:AC} describes how the
non-linear Allen-Cahn equation~\eqref{eq:SL} will be iteratively solved
using an IE formulation. Then, \S\ref{subsec:NumericalIssues} describes
how we will combine preconditioning strategies with fast summation
methods to accelerate the time to solution, and
\S\ref{subsec:timeStepping} describes adaptive time stepping methods.

>>>>

\subsection{Specific Aim 3: Efficient, high-order methods for
large-scale simulations}
\label{subsec:specific_aim_3}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 3, we propose novel and efficient high-order numerical methods to
simulate dynamic assembly of particles under the HAP in a viscous
solvent.

The hydrophobic interactions come from the tendency of particles to
minimize the free energy of the structure of the surrounding water
molecules. The HAP model uses the free energy
\begin{align}
\label{eq:free_energy}
F[u] = C \int_{\Omega} \rho |\nabla u|^2 + \rho^{-1} f(u) \,dx
\end{align}
where $u$ is an order parameter for the structure of water, $\rho$ is a
decay length, $C$ is a constant, and $f(u)$ is a potential.
Minimizing $F[u]$ requires that $u$ satisfies 
\begin{align}
  \label{eq:SL}
  -\rho^2 \Delta u + f'(u) = 0  \text{ in } \Omega,\quad u = g
  \text{ on } \Omega.
\end{align}
where the boundary condition $g$ is a material label that is transported
with the particle motions. The motion of the solvent phase is modeled by
the Stokes equations for an incompressible, viscous fluid, 
\begin{align}
  -\mu \Delta \uu + \nabla p = 0, \quad \nabla \cdot \uu = 0,
  \label{eq:stokes}
\end{align}
with vanishing conditions at infinity. The hydrodynamic and
hydrophobicity are copuled through viscous and hydrophobic stress
balance.

We will develop methods based on combinations of integral equation-based
volume and layer potentials, and inexact Newton's method to solve the
elliptic PDEs~\eqref{eq:SL} and~\eqref{eq:stokes}. Numerical methods to
solve these integral equations (IEs) have several advantages over their
PDE-based counterpart: by construction, they satisfy far-field and mass
conservation conditions; only $\partial\Omega$ needs to be discretized
for layer potentials; carefully chosen IEs result in a well-conditioned
linear system that is solved with a mesh-independent number of GMRES
iterations; and high-order or spectral accuracy is attainable with
appropriate quadrature methods. The PIs have demonstrated that IEs are a
powerful tool to simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}. Moreover,
elements of these formulations have been described in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016,
koh-cor-che-vee2021}. This proposal will extend the two-dimensional
methods to more general amphiphilic particle models that
satisfy~\eqref{eq:SL}, and all the methods can naturally be extended to
three dimensions. These methods will be motivated by models and analysis
developed in Specific Aims 1 and 2.  \S\ref{subsec:AC} describes how the
non-linear Allen-Cahn equation~\eqref{eq:SL} will be iteratively solved
using an IE formulation. Then, \S\ref{subsec:NumericalIssues} describes
how we will combine preconditioning strategies with fast summation
methods to accelerate the time to solution, and
\S\ref{subsec:timeStepping} describes adaptive time stepping methods.
